Illustrating Model Properties

### 2024-05-17 manuscript.qmd
add figure on ipsi & contra prevalence correlation

@@

In @fig-model-prevalences, we plot the prevalence of contralateral involvement of the LNLs II, III, and IV for the four scenarios made up of the possible combinations of early and late T-category, as well as lateralized and midline extending tumors.

{{< embed model.qmd#fig-model-prevalences >}}

@Fig-model-prevalences shows nicely how the model is capable of accurately taking the most important risk factors, i.e. T-category and midline extension, into account. As observed in the data, the model predicts that the prevalence of contralateral LNL II involvement jumps from below 8% for early T-category lateralized tumors to almost 40% when the tumor is of advanced T-category and crosses the mid-sagittal line.

@@

@@

>>>>

Note that we decided to omit the y-axis ticks and labels in these figures over prevalences and risks. The y-axis in these plots measures the probability density and its numerical values are not intuitively interpretable. Instead, we occasionally use the freed space to label e.g. rows of subplots.

@@

In @fig-model-prevalences-overall, we plot the prevalence of contralateral involvement of the LNLs II, III, and IV for the four scenarios made up of the possible combinations of early and late T-category, as well as lateralized and midline extending tumors.

{{< embed model.qmd#fig-model-prevalences-overall >}}

@Fig-model-prevalences-overall shows nicely how the model is capable of accurately taking the most important risk factors, i.e. T-category and midline extension, into account. As observed in the data, the model predicts that the prevalence of contralateral LNL II involvement jumps from below 8% for early T-category lateralized tumors to almost 40% when the tumor is of advanced T-category and crosses the mid-sagittal line.

@@

{{< embed model.qmd#fig-model-prevalences-with-ipsi >}}

In @fig-model-prevalences-with-ipsi we display the model's ability to capture the correlation between ipsi- and contralateral involvement. It shows that the prevalence of metastases in the two sides of the neck is correlated via the time of diagnosis, despite the model not having any direct connections between the two side. However, there are some small discrepancies in the model's prediction: When considering the scenario of LNL II involvement in the ipsi- *and* contralateral side, it cannot quite capture the split between early and late T-category. The model slightly overestimates the prevalence of this scenario for early T-category patients and underestimates it for advanced T-category.

## Prediction of Risk for Occult Disease

- compute risks for occult disease, given diagnosis

@@

# Discussion

- conceptually extended model to cover both sides of neck
- explicitly models midline extension (also allows for easy marginalization)
- midline extension intuitively causes contralateral spread to become more similar to ipsilateral spread

### 2024-05-17 model.qmd
add figure on ipsi & contra prevalence correlation

axes[0].set_ylim(0., 1.)
axes[0].yaxis.set_major_formatter(ticker.StrMethodFormatter("{x:.0%}"))
axes[0].set_ylabel("probability")
axes[0].grid(axis="y", color=COLORS["gray"], alpha=0.5)

@@

```{python}
#| echo: false
#| label: fig-model-burnin-history 
#| fig-cap: Monitoring quantities during the burn-in phase of the parameter sampling.
from collections import namedtuple
import pandas as pd

@@

```{python}
#| label: fig-model-state-dist
#| fig-cap: 3D state distribution of a midline model with 3 LNLs (II, III, and IV) in both sides of the neck.
from matplotlib.colors import LinearSegmentedColormap
from mpl_toolkits.axes_grid1 import AxesGrid

@@

To verify that the model does what it is supposed to, we compare how often some selected diagnoses appear in the data and how often the model predicts these diagnoses to be.

```{python}
#| label: fig-model-prevalences
#| fig-cap: Comparison of predicted (histograms) vs observed (beta posteriors) prevalences. Shown for the contralateral LNLs II (blue), III (orange), and IV (green). The top row shows scenarios with early T-category tumors, the bottom row for late T-category ones. The left column depicts scenarios where the primary tumor is clearly lateralized, the right column scenarios of tumors extending over the mid-sagittal line. 
from matplotlib import gridspec
from lyscripts.scenario import Scenario
from lyscripts.plot.utils import Histogram, BetaPosterior, draw
from scripts import paths

@@

full.set_title(
  "Observed vs Predicted Prevalence of Contralateral Involvement",
  pad=20.,
  fontsize="large",
)

for i, t_stage in enumerate(["early", "late"]):
  axes[i,0].set_ylabel(f"{t_stage} T-cat.", fontweight="bold")
  for j, midext in enumerate([False, True]):
    axes[0,j].set_title("Mid. ext." if midext else "Lateralized", fontweight="bold")
    axes[0,j].set_xticks([])
    axes[-1,j].set_xlabel("prevalence [%]")
    axes[i,j].set_yticks([])
    content = []
    for lnl, color in zip(
      ["II", "III", "IV"],
      [COLORS["blue"], COLORS["orange"], COLORS["green"]],
    ):
      scenario = Scenario(
        t_stages=[t_stage],
        midext=midext,
        diagnosis={"contra": {"max_llh": {lnl: True}}}
      )
      content.append(
        Histogram.from_hdf5(
          filename=paths.model_dir / "full" / "prevalences_overall.hdf5",
          dataname=scenario.md5_hash("prevalences"),
          color=color,
          label=f"contra LNL {lnl}"
        )
      )
      content.append(
        BetaPosterior.from_hdf5(
          filename=paths.model_dir / "full" / "prevalences_overall.hdf5",
          dataname=scenario.md5_hash("prevalences"),
          color=color,
        )
      )
    draw(
      axes=axes[i,j],
      contents=content,
      xlims=(0., 15. if not midext else 45.),
    )
    axes[i,j].legend(labelspacing=0.1)

@@

>>>>

axes[0].set_ylim(0., 1.)
axes[0].tick_params(bottom=False, labelbottom=False)
axes[0].yaxis.set_major_formatter(ticker.StrMethodFormatter("{x:.0%}"))
axes[0].set_ylabel("probability")
axes[0].grid(axis="y", color=COLORS["gray"], alpha=0.5)

@@

```{python}
#| echo: false
#| label: fig-model-burnin-history 
#| fig-cap: "Monitoring quantities during the burn-in phase of the parameter sampling."
from collections import namedtuple
import pandas as pd

@@

```{python}
#| label: fig-model-state-dist
#| fig-cap: "3D state distribution of a midline model with 3 LNLs (II, III, and IV) in both sides of the neck."
from matplotlib.colors import LinearSegmentedColormap
from mpl_toolkits.axes_grid1 import AxesGrid

@@

To verify that the model does what it is supposed to, we compare how often some selected diagnoses appear in the data and how often the model predicts these diagnoses to be observed.

## Contralteral LNLs II, III, and IV Dependent on T-Cat. and Mid. Ext.

We start witht the most commonly involved LNLs in the contralateral neck: II, III, and IV. We investigate their prevalence in the four scenarios early T-cat. withour mid. ext, late T-cat. without mid. ext., early T-cat. with mid. ext., and late T-cat. with mid. ext.

```{python}
#| label: fig-model-prevalences-overall
#| fig-cap: "Comparison of predicted (histograms) vs observed (beta posteriors) prevalences. Shown for the contralateral LNLs II (blue), III (orange), and IV (green). The top row shows scenarios with early T-category tumors, the bottom row for late T-category ones. The left column depicts scenarios where the primary tumor is clearly lateralized, the right column scenarios of tumors extending over the mid-sagittal line. This figure illustrates the model's ability to describe the prevalence of different combinations of scenarios involving the risk factors T-category and midline extension."
from matplotlib import gridspec
from lyscripts.scenario import Scenario
from lyscripts.plot.utils import Histogram, BetaPosterior, draw
from scripts import paths, shared

@@

full.set_title(
  "Observed vs Predicted Prevalence of Contralateral Involvement\nin the LNLs II, III, and IV",
  pad=20.,
  fontsize="large",
)

for i, t_stage in enumerate(["early", "late"]):
  axes[i,0].set_ylabel(f"{t_stage} T-cat.", fontweight="bold")
  for j, midext in enumerate([False, True]):
    axes[0,j].set_title("Mid. ext." if midext else "Lateralized", fontweight="bold")
    axes[0,j].set_xticks([])
    axes[-1,j].set_xlabel("prevalence [%]")
    axes[i,j].set_yticks([])
    content = []
    for lnl, color in zip(
      ["II", "III", "IV"],
      [COLORS["blue"], COLORS["orange"], COLORS["green"]],
    ):
      scenario = Scenario(
        t_stages=[t_stage],
        midext=midext,
        diagnosis={"contra": {"max_llh": {lnl: True}}}
      )
      content.append(
        Histogram.from_hdf5(
          filename=paths.model_dir / "full" / "prevalences_overall.hdf5",
          dataname=scenario.md5_hash("prevalences"),
          color=color,
          label="prediction",
        )
      )
      beta_post = BetaPosterior.from_hdf5(
        filename=paths.model_dir / "full" / "prevalences_overall.hdf5",
        dataname=scenario.md5_hash("prevalences"),
        color=color,
      )
      beta_post.kwargs["label"] = f"data: {beta_post.num_success} of {beta_post.num_total}"
      content.append(beta_post)
    draw(
      axes=axes[i,j],
      contents=content,
      xlims=(0., 15. if not midext else 45.),
    )

# manually place legends
shared.split_legends(
  ax=axes[0,0],
  titles=["LNL II", "LNL III", "LNL IV"],
  locs=[(0.57, 0.55), (0.48, 0.99), (0.15, 0.99)],
)
shared.split_legends(
  ax=axes[0,1],
  titles=["LNL II", "LNL III", "LNL IV"],
  locs=[(0.67, 0.99), (0.27, 0.6), (0.1, 0.99)],
)
shared.split_legends(
  ax=axes[1,0],
  titles=["LNL II", "LNL III", "LNL IV"],
  locs=[(0.64, 0.99), (0.27, 0.6), (0.24, 0.99)],
)
shared.split_legends(
  ax=axes[1,1],
  titles=["LNL II", "LNL III", "LNL IV"],
  locs=[(0.64, 0.99), (0.4, 0.6), (0.16, 0.99)],
)

@@

## Correlation between Ipsi- and Contralateral Involvement

```{python}
#| label: fig-model-prevalences-with-ipsi
#| fig-cap: "Comparison of predicted (histograms) vs observed (beta posteriors) prevalences. Shown are four scenarios, all including contralateral LNL II involvement: Early T-category and an ipsilateral N0 neck (green), early T-category and ipsilateral LNL II involvement (blue), as well as the same two scenarios but for advanced T-category (orange and red). The figure shows that the model is capable of describing the correlation between ipsi- and contralateral involvement. Although for the scenario of LNL II involvement in both sides, the prediction's split between early and advanced T-category is not large enough."
from lyscripts.scenario import Scenario
from lyscripts.plot.utils import Histogram, BetaPosterior, draw
from scripts import paths, shared

plt.rcParams.update(shared.get_fontsizes())
plt.rcParams.update(shared.get_figsizes(width=17./2.))

fig, ax = plt.subplots()
content = []
colors = iter([COLORS[c] for c in ["orange", "green", "red", "blue"]])

for t_stage in ["early", "late"]:
  for diagnosis in [{"II": True}, {lnl: False for lnl in ["I", "II", "III", "IV", "V"]}]:
    color = next(colors)
    scenario = Scenario(
      t_stages=[t_stage],
      diagnosis={
        "ipsi": {"max_llh": diagnosis},
        "contra": {"max_llh": {"II": True}},
      }
    )
    content.append(
      Histogram.from_hdf5(
        filename=paths.model_dir / "full" / "prevalences_with_ipsi.hdf5",
        dataname=scenario.md5_hash("prevalences"),
        color=color,
        label="prediction"
      )
    )
    beta_post = BetaPosterior.from_hdf5(
      filename=paths.model_dir / "full" / "prevalences_with_ipsi.hdf5",
      dataname=scenario.md5_hash("prevalences"),
      color=color,
    )
    beta_post.kwargs["label"] = f"data: {beta_post.num_success} of {beta_post.num_total}"
    content.append(beta_post)

draw(ax, content, xlims=(0., 25.))
ax.set_title("Observed vs Predicted Prevalence\nof Contralateral LNL II Involvement")
ax.set_yticks([])
ax.set_xlabel("prevalence [%]")

shared.split_legends(
  ax=ax,
  titles=["early with ipsi II", "early ipsi N0", "late with ipsi II", "late with ipsi N0"],
  locs=[(0.23, 0.65), (0.1, 0.99), (0.6, 0.65), (0.45, 0.99)],
)
